The 2nd moment of the Hitting time of an OU Process

### 2013-08-22 OU_HittingTimeVariance.tex
Initial commit

%%This is a very basic article template.
%%There is just one section and two subsections.
\documentclass{article}

@@

\usepackage{algpseudocode}
\usepackage{algorithm}

@@

\usepackage{cleveref}
\usepackage{cancel}

@@

\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Theorem}[thm]
% \theoremstyle{definition}
\newtheorem{ex}{Example}[thm]
\newtheorem{defn}{Definition}[thm]

@@

\title{Stochastic Optimal Control - with a view towards spike control in neurons
\\
{\small Advanced Comprehensive Exam in Mathematics}
\\
{\small Dept. of Math - University of Ottawa}
}
\author{Alexandre Iolov 
$<$\href{mailto:aiolo040@uottawa.ca}
		{aiolo040 at uottawa dot ca}$>$}

\date{January 10, 2013}

@@

\abstract{We review main concepts from stochastic optimal control theory,
including stochastic differential equations, dynamic programing and Pontryagin's
Maximum Principle. We then discuss an optimal control problem in a simple
mathematical model of a neuron where the feedback is minimal and we have a
quasi-open-loop control which necessitates the use of a Maximum Principle for
the system's transition density.}

\tableofcontents

\section{Overview}
Stochastic Optimal Control in continuous time deals with finding
the input which optimizes the performance of a continuous-time dynamical systems
that contain a significant random component in their evolutions.

Here we present a brief overview of the subject starting from the basic notions
of Stochastic Differential Equations (SDEs), 
and then introducing the main tools of Optimal Control, namely the
Hamilton-Jacobi-Bellman equation of Dynamic Programming and the state-adjoint
equations arising from Pontryagin's Maximum Principle. We will not state
proofs, although we sometimes provide heuristic derivations of the main
results.

If I had to learn the material all over again, I would proceed my readings as
follows: First I would read the online notes by Evans on both
SDEs,\cite{Evansa}, and Optimal Control, \cite{Evansb}. I would then read
Jacobs' book, \cite{Jacobs}, especially for the chapter on the Fokker-Planck
equation (ch 7) and then Oksendal \cite{Oksendal2007} for a
more thorough mathematical treatment of SDEs. Having familiarized myself with SDE's I would
return to Optimal Control in the Deterministic Case, starting with \cite{Kirk2004} for the
finite dimensional deterministic case (ODEs) and then \cite{Fleming1975} for the
extension to the stochastic case. I would keep the following for reference \cite{Krylov2008},
\cite{Hanson2007} and finally \cite{Fattorini1999} for the
deterministic infinite-dimensional (PDEs) case.

I have not used any particularly sophisticated numerical
methods, but, for example, Hanson \cite{Hanson2007} discusses both the
Euler-Maruyama scheme for simulating SDEs and the Crank-Nicholson Method for
solving PDEs which is all I have used in my codes.

All of the theoretical results presented in the sequel will be from
the above references.

\section{Stochastic Differential Equations}
In view of our ultimate goals, we will restrict ourselves to  stochastic
processes whose sample paths have continuous paths, i.e.\ to SDEs driven by
Brownian motion.

Since the reader is more likely well familiar with the material in this section,
we will not provide proofs, but only state the results with a view towards
establishing the notation for the sequel.

We will assume that the reader is familiar with the following concepts:
\begin{itemize} 
  \item a probability space, $\{\O, \sAlg, P\}$ consisting of a
  probability space, a sigma-algebra and a probability measure
  \item a continuous-time stochastic process, $X_t$
  \item a filtration $\Fil(t)$ and in particular the filtration generated by a
  stochastic process.
\end{itemize}

\subsection{The Wiener Process and the Ito Integral}
The Wiener Process is the fundamental building block of the stochastic calculus,
it is often called Brownian Motion and we will denote it $\{W_t\}_{t\geq 0}$. It
satisfies:
\begin{defn}Wiener Process, $W_t$:
\begin{enumerate}
  \item $W_0 = 0$
  \item $W_t - W_s = N(0, |t-s|)$ , i.e. normally distributed with mean 0,
  variance $|t-s|$)
  \item $\forall \{t_i\}_1^N, \quad \{W_{t_i} - W_{t_{i-1}} \}_2^N \sim$
  independent, i.e $W_t$ has independent increments
\end{enumerate}
\end{defn}
The fact that the finite incremental distributions suffice to specify a unique
continuous-time stochastic process is known as Kolmogorov's extension theorem.
\begin{ex}[from \cite{Evansa} \#21.] Inverted Wiener Process:
\\
Let $W_t$ be a 1-d Wiener Process and define:
\def \Wbar {{\bar{W} }}
\begin{equation*}
\Wbar(t) := \begin{cases}
t W(1/t) & t >0
\\
0 & t = 0
\end{cases}
\end{equation*}
Show $\Wbar(t)- \Wbar(s) \sim N(0, t-s)$.

\begin{proof}[Soln]
\begin{align*}
\Wbar(t) - \Wbar(s) =& t W(1/t) - s W(1/s), \, t>s
\\
=& (t-s)W(1/t) + s W(1/t) - sW(1/s) & \text{recall:} 1/s > 1/t
\\
=& (t-s)N(0, 1/t) - sN(0, 1/s -1/t) & \text{by indnt increments}
\\
=& N\left(0, \frac{(t-s)^2}t + \frac{s^2 }{ 1/s -1/t} \right) & \text{
addition of indnt normal RVs.}
\end{align*}
With some algebra, the variance in the last expression reduces to $t-s$
\end{proof}
\end{ex}

Now, we collect a few more relevant properties of $W_t$:
\begin{enumerate}
  \item The sample paths $W_{[0,\infty)}(\o)$
are almost surely (a.s.) continuous and are in fact Holder continuous for any
exponent $\g < 1/2$
\item The sample paths $W_{[0,\infty)}(\o)$ are nowhere differentiable
\item $W_t$ is a Markov process: $\Prob[W_t \in B | \s(W_{s' \leq s })] =
\Prob[W_t \in B \,| \,W_s]$, for any Borel set, $B \subset \R$, where $\s(W_{s'
\leq s })$ is the filtration generated by the $W_t$
\end{enumerate}

We now turn to defining stochastic integrals based on the Wiener Process.

\begin{defn} Progressively Measurable Functions:

Let $\Fil_t$ be the filtration generated by the Wiener Process.

Let $X_t$ be a stochastic process which is $\Fil_t$-measurable $\forall t$ and
which is jointly measurable in $(t,\o)$. We call such an $X_t$
\emph{progressively measurable}
\end{defn}

\begin{defn} $\Ltwopm, \Lonepm$

We define $\Ltwopm[0,T]$ as the space of all progressively measurable
$X$ such that
\begin{equation*}
\Exp \left[ \int_{[0,T]} X_t^2 \intd{t} \right] < \infty
\end{equation*}

Similarly, we define $\Lonepm[0,T]$ as the space of all progressively measurable
$X$ such that
\begin{equation*}
\Exp \left[ \int_{[0,T]} X_t \intd{t} \right] < \infty
\end{equation*} 
\end{defn}

$\Ltwopm$ will be the class of functions for which the Ito integral is
well-defined. 
\begin{defn} Ito Integral:
\label{defn:ito_integral}

Let $P^n := {a = t^n_1 \ldots t^n_{m_n} = b}$ be a partition of the interval
$[a,b] \subset [0, \infty)$. Let $|P_n| = \sup_i|t_i - t_{i-1}|$. 
Take $|P_n| \rightarrow_n 0 $ and consider an $ X_t \in \Ltwopm[a,b]$,

then
\begin{equation}
\int_{[a,b]} X_t \intd{W} := \lim_{n \rightarrow \infty}  
\sum_{i=1}^{m_n} X_{t_i}\left( W(t_{i+1}) - W(t_{i})\right)
\end{equation}

\end{defn}

To be precise, our definition is actually a theorem, and the real
definition is one that uses step functions and passes to the limit.  Also we
will write the limits of integration $\int_0^T \cdot  \intd{W}$ or $\int_{[0,T]}
\cdot  \intd{W}$ interchangeably.
% To justify def'n \ref{defn:ito_integral}, we recall:
% \begin{lemma}[Quadratic Variation]
% \begin{equation}
% \lim_{n \rightarrow \infty} \sum_{i=1}^{m_n} \left( W(t_i) - W(t_{i-1})\right)^2
% = b-a \quad \text{in }  L^2(\O)
% \end{equation}
% \end{lemma}
% The above can be read, heuristically as $\lim \sum (\Delta W)^2 \rightarrow \sum
% \Delta t$, which is the meaning behind the colloquial $\intd{W} =
% \sqrt{dt}$.

Again, we state without proof a few interesting properties of $\int X \intd{W}$:
\begin{thm} Ito Integral Properties

\begin{enumerate}
  \item $\Exp[\int_0^T X \intd{W} ] = 0$ 
  \item $\Exp[\left(\int_0^T X \intd{W}\right)^2 ] = \Exp[\int_0^T X^2
  \intd{t}]$
  \item $I(t) = \int_0^t X \intd{W} $ is a martingale 
  \item $I(t) = \int_0^t X \intd{W} $ has continuous sample paths
\end{enumerate}
\end{thm}

\subsection{Ito SDEs and Ito's Lemma}
\begin{defn} an Ito SDE:

We write
\begin{equation}
dX_t = Fdt + G dW
\end{equation}
on $0 \leq t \leq T$, if $X$ is a real-valued stochastic process satisfying:
\begin{equation*}
X(r) = X(s) + \int_{[s,r]} F \intd{t} + \int_{[s,r]} G \intd{W}
\quad 0\leq s \leq r \leq T
\end{equation*}
for some $F \in \Lonepm[0,T]$ and $G \in \Ltwopm[0,T]$
\end{defn}

We are now ready to present the celebrated Ito Lemma which is the chain-rule of
stochastic calculus:

\begin{thm}[Ito Lemma] Suppose $X$ satisfies the stochastic differential $dX_t =
Fdt + G dW$ as above and take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$.

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)
\intd{t} + \left(   \di_x u\cdot G  \right)\intd{W}
$$
\end{thm}

% We now present two examples, one fundamental and one more involved:
% \begin{ex}[pg 67 in \cite{Evansa}] The 'Stochastic' Exponential:
% 
% Let $Y = \exp(\l W - \l^2 t / 2)$. In the notation of the above Thm, we have:
% $dX = dW, F=0, G=1$ and $u(x,t) = \exp(\l x - \l^2 t / 2)$.
% 
% then
% \begin{align*}
% dY =& 
% \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)\intd{t} +
% \left(   \di_x u \cdot G  \right)\intd{W}
% \\
% =&
% \left( -\frac{\l^2 t}{2}\exp(\l x - \l^2 t / 2) + \l^2\exp(\l x - \l^2 t / 2)  \right)\intd{t}
% \\ 
% &+ \left(   \l \exp(\l x - \l^2 t / 2) \right)\intd{W}  
% \\
% =&  \l \left( \exp(\l x - \l^2 t / 2) \right)\intd{W}  
% \\
% =& \l Y dW
% \end{align*}
% \end{ex}
% 
% \begin{ex}[pg. 67 in \cite{Evansa}] Hermite Polynomials:
% 
% Let $h_n(x,t) = \frac{(-t)^n}{n!} e^{x^2/2t} \di^n_x \left( e^{x^2/2t} 
% \right)$ be the $n$th Hermite polynomial
% \\
% then
% $$
% \int_0^t h_n(W, t) \intd{W} = h_{n+1}(W(t), t)) 
% $$
% i.e. in Ito calculus, the Hermite polynomials are like the power monomials
% $\frac{t^n}{n!}$ in the ordinary calculus.
% \begin{proof} Since
% $$
% d_\l^n [\exp(- \frac{(x-\l t)^2}{2t} ] |_{\l=0} =
%  (-t)^n d_x^n [ \exp(-\frac{x^2}{2t})] $$
% We will have:
% \begin{align*}
% d_\l^n [\exp(\l x - \l^2 t / 2) ] |_{\l=0} &= (-t)^n  e^{x^2/2t} d_x^n [
% \exp(-\frac{x^2}{2t})]
% \\
% &= n! h_n(x,t)
% \end{align*}
% Thus we can expand $\exp(\l x - \l^2 t / 2) $ in a Taylor series at $\l=0$ as:
% $$
% \exp(\l x - \l^2 t / 2) = \sum_n \l^n h_n(x,t)
% $$
% Now recall $Y$ from the previous example:
% \begin{equation*}
% Y(t) = \exp(\l W(t) - \frac{\l^2 t }2) = \sum_n \l^n h_n(W(t),t)
% \end{equation*}
% which satisfies:
% \begin{equation*}
% Y(t) = 1 + \l \int Y \intd{W}
% \end{equation*}
% Naturally we will plug the series expansion in the SDE to obtain:
% \begin{align*}
% \sum_{n=0}^{\infty} \l^n h_n(x,t) =&
% 1 + \l \int_0^t  \sum_{n=0}^{\infty} \l^n
% h_n(x,t)\intd{W}
% \\
% =& 1 + \sum_{n=1}^{\infty} \l^n\int_0^t 
% h_{n-1}(x,t)\intd{W} \quad {\textrm{bump $n$ up by one}}
% \end{align*}
% Equating the coefficients for $\l^n$ on both sides leads to the result
% \end{proof}
% \end{ex}

Finally, we are in a position to define an Ito SDE:

\begin{defn}[Ito SDE] Let
$F(x,t):\R\times [0,T]\ra \R,  G(x,t):\R\times [0,T]\ra \R,$ be given functions.
We say that a stochastic process $X$ satisfies:
\begin{equation}
dX =F dt + G dW
\end{equation} 
over $[0,T]$ if
\begin{enumerate}
  \item $X$ is progressively measurable wrt. $\Fil_t$
  \item $F(X,t) \in \Lonepm$
  \item $G(X,t) \in \Ltwopm$ 
  \item $X_t = X_0 + \int_0^t F(X_s, s) \intd{s} + \int_0^t G(X_s, s) \intd{W_s}$
\end{enumerate}
\end{defn}

% \begin{ex} The solution to 
% $$
% \begin{cases}
% dX = F(t)X dt + G(t) X dW
% \\
% X_0 = 1
% \end{cases}
% $$
% is 
% $$
% X_t = \exp\left( \int_0^t F(s) - \frac{G^2(s)}{2} \intd{s} + \int_0^t G
% \intd{W} \right) $$
% \begin{proof}
% Let $Y = \int_0^t f(s) - \frac{g^2(s)}{2} \intd{s} + \int_0^t g
% \intd{W} $
% and guess that $X = e^Y$
% then, using Ito's lemma:
% \begin{align*}
% dX =& (\di_y [e^y] dY + \frac{1}{2}\di^2_y[e^y]  g^2 dt
% \\ =&
% (e^y  f dt - e^y g^2 / 2) dt + e^y g dW + \frac{1}{2} e^y \cdot  g^2 dt
% \\ =& e^y \left( f dt +g dW \right)
% \\ =& X \left( f dt +g dW \right)
% \end{align*}
% \end{proof}
% \end{ex}

% \begin{ex}[Brownian Bridge, pg 80 in \cite{Evansa}] The solution of the SDE
% $$
% \begin{cases}
% dB = -\frac{B}{1-t} dt + dW
% \\
% B_0= 0
% \end{cases}
% $$
% is
% $$
% B(t) = (1-t) \int_0^t \frac{1}{1-s} dW
% $$
% \begin{proof}
% It is not clear how to apply Ito's lemma to this problem, so instead we just
% directly calculate the differential:
% \begin{align*}
% B(t^+) - B(t) =& (1-t^+) \int_0^{t^+} \frac{1}{1-s} dW - (1-t) \int_0^t
% \frac{1}{1-s} dW
% \\ =&
% (1-t^+) \int_t^{t^+}\frac{1}{1-s} dW  - (t^+-t) \int_0^t
% \frac{1}{1-s} dW
% \\ \approx&
% \int_t^{t^+} dW - \Delta t \frac{B(t)}{1-t}  
% \end{align*}
% \end{proof}
% \end{ex}
% Actually we can say a bit more about the Brownian bridge, in particular we
% will calculate $\Exp[B^2]$
%  \begin{ex}[Evans .37]
% \begin{align*}
% \Exp[B^2] =& \Exp\left[\left( (1-t) \int_0^t \frac{1}{1-s} dW \right)^2 \right]
% \\
% =&(1-t)^2 \Exp\left[ \int_0^t\left(  \frac{1}{1-s}\right)^2 \intd{s} \right]
% \\
% =&(1-t)^2 [ \frac{1}{1-t} - 1 ]
% \\
% =&(1-t) - (1-t)^2
% \end{align*}
% 
% Further note that  $\lim_{t\ra 1^-}\Exp[B^2] = 0$ and that $B$ is a martingale
% and so its first moment is always equal to its initial value, $0$. So indeed the 2nd
% moment is the variance and with the variance converging to zero, we can use
% Chebyshev's inequality to conclude that $\lim_{t\ra 1^-}B(t) = 0 $(a.s). This
% justifies the name of the Brownian bridge - it is a normally distributed RV.
% which is clamped to equal 0 at both $t= \{0,1\}$.
% \end{ex}

As an example we will discuss the Ornstein-Uhlenbeck process, which is the
basis for the models we will face later on.
\begin{ex}[O-U Process] Let $X$ follow:
\begin{equation*}
dX = \th (\a - X_t) dt + \b dW
\end{equation*}
with an initial condition, $X_0$, which may be an arbitrary distribution
independent of the Wiener Process,$W$.
We can solve this as follows:
\begin{align*}
dX =& \th (\a - X_t) dt + \b dW
\\
dX + \th X_t) dt=& \th \a dt + \b dW
\\
e^{\th t} dX + e^{\th t}\th X_t) dt=& \th e^{\th t}\a dt + \b e^{\th t} dW
\\
Xe^{\th t} - X_0 = & \int \th e^{\th t}\a dt +  \int \b e^{\th t} dW
\\
X_t =& e^{-\th t} X_0 + \a(1-e^{-\th t}) +  \frac{\b e^{-\th t}}{\sqrt{2\th}}
W(e^{2\th t}-1)
\end{align*}
which means that $X$ forgets its initial conditions exponentially fast and
converges to a normal random variable with mean $\a$ and variance
$\tfrac{\b^2}{2\th}$
\end{ex}

At the end of this sub-section, we state Ito's Lemma with multiple Wiener
processes and then for a multidimensional state.

\begin{thm}[Ito Lemma for $dW \in  \R^{m}$] Suppose $X$ satisfies the stochastic
differential $dX_t = F dt + G dW$, where $ F \in \R, G \in \R^{1\times m}$ and
$dW = \left(dW^{(i)}\right) \in  \R^{m}$ is a vector of independent Wiener
Processes. Take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$. Let $D :=\left[{\sum_k
G_{k} G_{k} } \right] \in \R$

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u +  \di_{x} u \cdot F + 
  \di^2_{x} u \cdot D \right)
\intd{t} +
 \left(  \di_{x} u  \cdot  G \right) dW  
$$
\end{thm}

\begin{thm}[Ito Lemma for $X \in R^n$]
Suppose $X$ satisfies the stochastic differential $dX_t = F dt + G dW$, where $
F \in \R^n, G \in \R^{n\times m}$ and $dW = \left(dW^{(i)}\right) \in  \R^{ 
m}$ is a vector of independent Wiener Processes. Take $u(x,t) \in C^{2,1}[ \R^n
\times [0,T]]$. Let $D_{ij} := \tfrac{1}{2} \left[{\sum_k G_{ik} G_{jk} }
\right]_{ij} \in \R^{n\times n} $

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u + \sum_i \di_{x_i} u \cdot f_i + 
\sum_{i,j} \di^2_{x_i x_j} u \cdot D_{ij} \right)
\intd{t} +
 \left(  \sum_{ij} \di_{x_i} u  \cdot G_{ij} dW^{(j)} 
\right)$$
\end{thm}

\subsection{Fokker-Planck and Kolmogorov's Backward Equations}
The Fokker-Planck equation and Kolmogorov's Backward equation describe the
forward (resp. backward)  evolution of the probability density of $X_t$

For the rest of this section we will work in $ \R^n$, i.e. $X$ will be an $n-$
dimensional vector that satisfies the stochastic differential 
\begin{equation}
dX_t = F(X,t) dt + G(X,t) dW,
\label{eq:generic_Ito_SDE_Rn}
\end{equation}
$F \in \R^n, G \in \R^{n \times m}$,
$dW = (dW^{(i)}) \in  \R^{  m}$ is an $m$-dimensional Brownian motion and we
write $D_{ij} :=
\tfrac{1}{2} 
\left[{\sum_k G_{ik} G_{jk} } \right]_{ij} \in \R^{n\times n} $.

% Let $$\F(x,t| y,s) = \Prob[X_t \preceq x| X_s = y] $$ be the forward transition
% distribution and

Let $$ \f(x,t| y,s) \intd{x} =  \Prob[X_t \in \intd{x} | X_s = y] $$ be the
transition probability density. Then $\f$ satisfies:
\begin{equation}
\di_t \f= -\sum_i \di_{x_i} \left[ F_i(x,t) \f(x,t) \right] 
+ 
\sum_{i,j}  \di^2_{x_i x_j} \left[ D_{ij}(x,t) \f(x,t) \right]
\label{eq:FP_pde}
\end{equation}
This is called the \emph{Fokker-Planck} or \emph{Forward Kolmogorov} equation.
It can also be seen as a continuity euquation or a cosevation of probability
equation. To this end define the probability current, $\p \in \R^n$ as:
$$
\phi_i =   F_i(x,t) \f(x,t) 
+ 
\sum_{j}  \di_{x_j} \left[ D_{ij}(x,t) \f(x,t)\right]
$$
then the Fokker-Planck equation, \cref{eq:FP_pde}, can be written as:
$$
\di_t \f= -\grad \cdot \phi
$$
which just says that the change in probability is the
difference between the flow in and the flow out.

Initial conditions for $\f$ are given by the distribution of $X_s$, if $s=0$
then this is the initial distribution of $X$. Boundary conditions (BCs) depend
on the domain of $X$ and what happens to $X$ once it hits its boundary. If the domain
is all of $\R^n$ then we only insist that $\lim_{|x| \ra \infty} \f = 0$. If the
domain has boundaries, then there are two common scenarios which we will also
encounter in the sequel:
\begin{enumerate}  
  \item absorbing BCs
  \item reflecting BCs
\end{enumerate}
At an absorbing boundary, the particle $X$ is removed and $\f=0$ there.
In this situation, we will not have conservation of probability and the integral
of $\f$ over $X$'s domain will monotonically decrease.

At a reflecting boundary, the particle $X$ bounces back into its domain and we
will have that the probability current $\p\cdot n = 0$ where $n$ is the outward
normal at the reflecting boundary.

We now consider $\f$ as functions of $y,s$, holding $x,t$ fixed, then
\begin{equation}
-\di_s \f= \sum_i  \left[F_i(y,s) \cdot \di_{y_i}\f(|y,s) \right] 
+ 
\sum_{i,j}   \left[ D_{ij}(x,t) \di^2_{y_i y_j}\f(|y,s) \right]
\label{eq:FP_backward_pde}
\end{equation} 
This is the \emph{Backward Kolmogorov} equation for $\f$. Note the minus sign
in front $\di_s \f$. A mnemonic for the signs of the Backward vs. Forward
equation is that as $t$ increases $\f$ diffuses and so $\di_t$ and $\di^2_{x}$
have the same sign, but as $s$ increases $\f$ anti-diffuses and they have
opposite signs.

The differential operator on the right-hand side of
\cref{eq:FP_backward_pde} occurs often in the study of SDEs and has its own
name.
\begin{defn}[Generator of an SDE] the Generator $A$ of $X$ is defined by:
$$
A[\psi(x)] = \lim_{t \searrow 0^+} \frac{\Exp[\psi(X_t)]  - \psi(x)}{t} ;  \quad X_0 =
x \in \R^n
$$
\end{defn} 
\begin{lemma} For an Ito SDE as in \cref{eq:generic_Ito_SDE_Rn} 
$$
A[\psi(x)] = \sum_i F_i(x,t) \cdot \di_{x_i} \psi + \sum_{i,j} D_{ij}(x,t)
\di^2_{x_i x_j}\psi $$
\end{lemma}

\subsection{Stopping Times}
\begin{defn}
Let $\Fil(t)$ be some filtration, a random variable $\t$ is called a
\emph{stopping time} if 
$$
\{\o : \t(\o) \leq t\} \in \Fil(t) \, \forall t  
$$
\end{defn}
The colloquial way of describing stopping times is at any time we know
whether $\t$ has occurred or not. For a counterexample, the time that a Wiener
Process achieves its maximum over some interval is not a stopping time, since at any
given time, we do not know if the maximum has occurred or not. The most common
example of a stopping time is the first hitting-time, which is the first time
$X_t$ leaves a set. 
\begin{thm}[First-Hitting time] Let $E \subset \R^n, E \neq
\phi$ be open or closed

then $$\t := \inf\{ t \geq 0 | X_t \in E\}$$ is a stopping time.
\end{thm}

The reason stopping times are very useful is that all the facts so far quoted
for Ito calculus using integrals $\int_0^T \intd{W}$ remain true if $T$ is
replaced by a stopping time $\t$.

Also it allows us to link SDE's and PDEs using the generator, $A$, of the
diffusion: 
\begin{thm}[Dynkin's formula] Given $\t$ a stopping time, $\Exp[\t] < \infty$.
Then:
$$
\Exp[u(X_\t, \t)] =
u(x, 0) + \Exp\left[\int_0^\t \di_t u + A[u] \intd{s}
\right]
 $$
\end{thm}
This provides a link between PDEs and stochastic processes and allows us to go
back and forth in that we can find probabilistic results by
solving a PDE or we can approximate a PDE by simulating a stochastic process
and averaging.

\begin{ex}[\cite{Evansb} pg 99 - Expected hitting time to a boundary]
\label{ex:mean_hitting_time}
 Let $\O
\subset \R^n$ be a bounded open set with smooth boundary $\di \O$ then it is a
basic fact from PDEs theory that
\begin{equation}
\begin{cases}
-\frac{1}{2} \grad^2 u = 1  & \text{over } \O
\\
u =  0 &\text{on } \di \O
\end{cases}
\end{equation}
has a unique $C^\infty(\O)$ solution.

Let $X = W_t + x$ for any $x \in \O$ and define
 $$\t_x := \text{first time } X \text{ hits } \di \O$$
then the generator of $X$ is $A[\psi] = -\grad^2(\psi)/2$ and we will
have:

\begin{align*}
\Exp[u(X_\t)] - \Exp[ u(x_0)] =& \Exp \left[ -\frac{1}{2}\int_0^\t \grad^2
u\right]
\\
=& \Exp \left[ - \int_0^\t 1
\right]
\\
=-&\Exp [\t]
\end{align*}
Finally, invoke $u$'s BCs, $u|_{\di E} = 0$, and $X$'s ICs, $X_0=0 = x$ to
conclude that $$ u(x) = \Exp [\t]$$
The solution to the PDE evaluated at $x$ is the expected exit time from
$E$ for an $X_t$  starting at $x$.
\end{ex}
% Similarly we can calculate more moments of the hitting time.
% \begin{ex}[Mean squared hitting time to a boundary] Let $E
% \label{ex:mean_squared_hitting_time}
%  \subset \R$ be a bounded open set and let $u$ solve:
%  \begin{equation}
%  \begin{cases}
%  (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2 u = -\frac t 2  &
%  \text{over } E
%  \\
%  u =  0 &\text{on } \di E
%  \end{cases}
%  \end{equation}
% 
% Then $- t / 2 = A[u]$ for $A$, the generator, corresponding to $dX = (\a - x /
% \tc) dt + \b dW$.
% Define $$\t_x := \text{first time } X \text{ leaves }  E$$
% then
%  \begin{align*}
%  \Exp[u(X_\t)] - \Exp[ u(X_0)] =& 
%  \Exp \left[  \int_0^\t   (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2
%  u \right]
%  \\
%  =& \Exp \left[ - \int_0^\t -\frac t 2 
%  \right]
%  \\
%  =-&\Exp [\t^2]
%  \end{align*}
% And we can conclude that 
% $$ u(x) = \Exp [\t^2]$$
% The solution to the PDE evaluated at $x$ is the expected squared exit time from
% $E$ for an $X_t$  starting at $x$.
% \end{ex}

>>>>

\documentclass{article}

@@

@@

\usepackage{cleveref}

@@

@@

\title{The 2nd moment of the Hitting time of an OU Process} 
\author{Alexandre Iolov 
$<$\href{mailto:aiolo040@uottawa.ca}
		{aiolo040 at uottawa dot ca}$>$}

\date{\today}

@@

\abstract{Calculate the 2nd moment of the hitting time of an OU process given
an arbitrary initial position}

\def \t {{ T}}

\section{Problem Formulation}
Given the following OU Model:
\begin{equation}
\begin{gathered}
dX_s = (\a - \frac{X_s}{\tc} ) \intd{s} + \b \intd{W_s},
\\
X(0) = x_0,
\\
\a > 0
\end{gathered}
\label{eq:X_evolution_uo}
\end{equation}
and a hitting time, $\t$, defined by
\begin{equation}
X_\t = \xth 
\quad \textrm{AND} \quad
X_{s < \t} < \xth
\end{equation}
we want an analytical expression for:
\begin{equation}
\Exp[\t^2|x_0]
\end{equation}
as a function of $\a,\tc,\b$

We can assume $\xth = 1.0$, but $x_0$ is arbitrary, that is - it is not
necessarily the 'equilibirum' value $.0$, indeed we will need to know
$\Exp[\t^2|x_0]$ for a whole range of $x_0$.

\subsection{Analytical Expressions for $\Exp[\t^2|x_0]$}

We will use the formulas/notation directly from \cite{Inoue1995}.
\begin{equation}
\Exp[\t^2] = \tc^2 \left[
2\phi_1^2(\eta) -
  \phi_2(\eta) -
  2\phi_1(\eta)\phi_1(\xi)+
  \phi_2(\xi)
  \right]
\end{equation}
where
\begin{eqnarray}
\xi &=& \boxed{(x_0 - \a\tc) \sqrt{\frac{2}{\b^2 \tc}}}
\\
\eta &=&
\boxed{(\xth -\a \tc )\sqrt{\frac{2}{\b^2 \tc}}}
\\
\phi_1(z)&=&
 \frac{1}{2} \sum_{n=1}^{\infty} \frac{(z \sqrt{2})^n}{n!}
 \Gamma\left(\tfrac{n}{2}\right)
\\
\phi_2(z)&=& 
 \frac{1}{2} \sum_{n=1}^{\infty} \frac{(z \sqrt{2})^n}{n!}
 \Gamma\left(\tfrac{n}{2}\right) \left( \psi\left(\tfrac{n}{2} \right) -
 \psi(1) \right)
\end{eqnarray}
where $\Gamma, \psi$ are the
\href{http://docs.scipy.org/doc/scipy/reference/generated/scipy.special.gamma.html#scipy.special.gamma}{Gamma}
and
\href{http://docs.scipy.org/doc/scipy/reference/generated/scipy.special.psi.html#scipy.special.psi}{digamma}
 functions respectively. Note that $\xth > x_0 \implies \eta > \xi$.

In the paper \cite{Inoue1995}, we are cautioned that evaluation of $\phi_k(z)$
directly is efficient and accurate as long as $z$ is positive and small. If $z$
is negative we will have an alternating series and a cancellation might cause
numerical errors.

Actually, I cannot get this to work at all!!!

% Instead we will use the Backward
% Kolmogorov equation, which is more useful since we need to calculate
% $\Exp[T^2|x_0]$ for a whole interval of $x_0$'s.

\subsection{Backward Kolmogorov equation}

Here we follow Jacobs, \cite{Jacobs}, in deriving and solving a differential
equation for $\Ttwo = \Exp[\t^2|x_0]$.
